---
title: "RunTerraWorkflow: Run workflows implemented in Terra"
author: "Sehyun Oh"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Run bioBakey using RunTerraWorkflow package}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    number_sections: yes
    toc: yes
    toc_depth: 4
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  comment = "#>", collapse = TRUE, fig.align = 'center'
)
```



# Overview
For R users with the limited computing resources, we introduce the 
RunTerraWorkflow package. This package allows users to run workflows 
implemented in [Terra](https://app.terra.bio/#), a cloud-based genomics 
platform, without writing any workflow, installing software, or managing cloud 
resources. Terra's computing resources rely on Google Cloud Platform (GCP) and 
to use RunTerraWorkflow, you only need to setup the Terra account once at the 
beginning.



## Create Terra account  
You need [Terra account setup](https://support.terra.bio/hc/en-us/articles/360034677651-Account-setup-and-exploring-Terra). Once you have your own Terra account, you need 
two pieces of information from it to use RunTerraWorkflow package: 
1) the email address linked to your Terra account and 2) your billing 
project name. Here are the basic input parameters used in this vignette. 
**Modify these with YOUR account information!**

```{r}
accountEmail <- "shbrief@gmail.com"
billingProjectName <- "waldronlab-terra-rstudio"
```


## Major steps 
Here is the table of functions for three major steps, prepare/run/check 
result, of the workflow.

| Steps | Functions | Description |
|--------|-----------|-------------|
| Prepare | `cloneWorkspace` | copy the template workspace |
|          | `updateInput` | take user's inputs |
| Run | `launchWorkflow` | launch the workflow in Terra |
|          | `abortSubmission` | abort the submission |
| Result | `monitorSubmission` | monitor the status of your workflow run |
|          | `listOutput` | display the list of your workflow outputs |
|          | `getOutput` | download your output |

## Load packages
```{r eval=TRUE}
if (!"RunTerraWorkflow" %in% installed.packages())
    devtools::install_github("shbrief/RunTerraWorkflow")
if (!"biobakeR" %in% installed.packages())
    devtools::install_github("shbrief/biobakeR")

suppressPackageStartupMessages({
  library(RunTerraWorkflow)
})
```

## Example: microbiome analysis
In this vignette, we are running [Whole Metagenome Shotgun (wmgx) analysis workflow](https://github.com/biobakery/biobakery_workflows) using 
[bioBakery](https://github.com/biobakery/biobakery/wiki) tools. You can find 
the currently available workspaces using `availableAnalysis()` function and 
the values under `analysis` column can be used for the analysis argument. For 
this vignette, we use `"bioBakery"`. 

```{r eval=TRUE}
availableAnalysis()

analysis <- "bioBakery"
```



# Setup
## Clone Workspace
First, you should clone the template workspace using `cloneWorkspace` function. 
Note that you need to provide a unique name for the cloned workspace through 
`workspaceName` argument.    

```{r}
workspaceName <- "mtx_workflow_biobakery_ver3"
```

If you attempt to clone the template workspace using the existing 
workspaceName, you will get the below error message.

```{r}
cloneWorkspace(accountEmail, billingProjectName, workspaceName, analysis)
```

With the unique workspace name, you can successfully clone the workspace and 
the function will return the name of the cloned workspace.

```{r}
cloneWorkspace(accountEmail, 
               billingProjectName, 
               workspaceName = "microbiome", 
               analysis)
```

```{r echo=FALSE, message=FALSE}
## Delete workspace
resp <- AnVIL::Terra()$deleteWorkspace(workspaceNamespace = billingProjectName,
                                       workspaceName = "microbiome")
rm(resp)
```


## Prepare Input
### Current input
You can review the current inputs using `currentInput` function. Below shows all
the required and optional inputs for the workflow.

```{r}
currentInput(accountEmail, billingProjectName, workspaceName)
```

<br>

`biobakeR::biobakery_currentInput` is a variation of `currentInput` function 
optimized for bioBakery workflow. It displays only the two major inputs - 
`inputListPath` and `inputFilePath`. These two types of files should be stored 
in Google bucket.   

1) `inputFilePath` file contains the paths to the input fastq files     
2) `inputListPath` file is a list of their paths as a txt file

In other words, `inputFilePath` is the content of `inputListPath` - you save 
fastq files and provide their file paths in the txt file.

This vignette uses six fastq files for the test. Using preemptive instances, 
this demo set will cost about $5 to run and take about 10min.

```{r}
biobakeR::biobakery_currentInput(accountEmail, 
                                 billingProjectName, 
                                 workspaceName)
```


### Update input
Before launching the workflow, you should provide the correct input information 
using `updateInput` function. RunTerraWorkflow doesn't support this function 
yet, but the bioBakery-specific function is available through `biobakeR::biobakery_updateInput`.

We use the six demo input files and their metadata.

```{r}
input <- "gs://run_terra_workflow/IBDMDB/ibdmdb_file_list.txt"
inputMeta <- "gs://run_terra_workflow/IBDMDB/ibdmdb_demo_metadata.txt"

biobakeR::biobakery_updateInput(accountEmail, billingProjectName, workspaceName,
                                InputRead1Files = input,
                                InputMetadataFile = inputMeta,
                                AdapterType = "NexteraPE",
                                ProjectName = "ibdmdb_test",
                                InputExtension = ".fastq.gz",
                                InputRead1Identifier = "_R1",
                                InputRead2Identifier = "_R2")
```



# Run bioBakery workflow
Once you clone the template workspace and update the input with your own 
data, you can launch the workflow using `launchWorkflow` function.

```{r}
launchWorkflow(accountEmail, billingProjectName, workspaceName)
```

## Monitor Progress
The last three columns show the submission and the result status.

```{r}
submissions <- monitorSubmission(accountEmail, 
                                 billingProjectName, 
                                 workspaceName)
submissions
```

## Abort submission
To abort the most recently submitted job, you don't need to specify 
`submissionId`.

```{r}
abortSubmission(accountEmail, billingProjectName, workspaceName)
```

If the submitted job is already done, you will get the below message.
```{r}
abortSubmission(accountEmail, billingProjectName, workspaceName,
                submissionId = "e20588f6-6bd9-4630-9cb1-de96aa9477f7")
```

If the submitted job is already aborted before done, you will get 
the below message.
```{r}
abortSubmission(accountEmail, billingProjectName, workspaceName,
                submissionId = "21840722-772c-4031-ab74-b7ae44f5e6f4")
```


# Result
## List Outputs 
You can check all the output files from the most recently succeeded submission.
If you specify the `submissionId` argument, you can get the output files of that
specific submission. Output files can also be subset using `keyword` argument.

```{r}
listOutput(accountEmail, billingProjectName, workspaceName)
```

```{r}
listOutput(accountEmail, billingProjectName, workspaceName, keyword = "humann")
```


```{r echo=FALSE, eval=FALSE}
## Find a specific submissionId
jobs <- RunTerraWorkflow::monitorSubmission(accountEmail, 
                                            billingProjectName, 
                                            workspaceName)
jobs_succeed <- which(jobs$succeeded == 1)
jobs[jobs_succeed,]

submission_id <- jobs[jobs_succeed,]$submissionId[1]
submission_id
```

If your outputs include tsv files, you can check the head of those files 
using `tableHead` function without downloading them. 

```{r warning=FALSE}
tableHead("HSM7J4NY_genefamilies_relab.tsv", n = 6,
          accountEmail, billingProjectName, workspaceName)   
```


## Get Outputs
`keyword` argument takes a character string containing a regular expression. In
the example below, we check all the `.tsv` outputs of the sample, "HSM7J4NY".

```{r}
listOutput(accountEmail, billingProjectName, workspaceName, 
           keyword = "HSM7J4NY.*.tsv")
```

You can download any file using `getOutput` function. Here, we narrow down the 
download to HSM7J4NY sample's `.tsv` output files.

```{r echo=FALSE}
HSM7J4NY_dir <- "~/data2/RunTerraWorkflow/inst/extdata/outputs/HSM7J4NY/"
```

```{r eval=FALSE}
getOutput(accountEmail, billingProjectName, workspaceName, 
          keyword = "HSM7J4NY.*.tsv", dest_dir = HSM7J4NY_dir)
```

```{r}
list.files(HSM7J4NY_dir)
```

