---
title: "Create search functions for Terra workspace, workflow, and sample queries"
author:
  - Sehyun Oh
  - Kai Gravel-Pucillo
date: "`r format(Sys.time(), '%B %d, %Y')`"
format:
    html:
        fontsize: 14pxs
        toc: true
        top-depth: 3
abstract: "Prepare U24 Supplement: AI/ML-ready"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                    warning = FALSE,
                    message = FALSE,
                    collapse = TRUE,
                    eval = TRUE)
```

# Overview

This .qmd file demonstrates the use of the `workspace_search()` and `workflow_search()`
functions. These functions utilize the AnVIL package and are designed to query the 
Terra API.


# Setup

## Load Packages

```{r load}
suppressPackageStartupMessages({
  library(tidyverse)
  library(rols)
  library(dplyr)
  library(hash)
  library(vctrs)
  library(rvest)
  library(RSQLite)
  library(DBI)
  library(AnVIL)
})
```

## Data Setup

First we can load the tables containing the workspaces and workflows metadata:
```{r curation_setup, eval=FALSE}
# Specifying directory path: "C:/users/owner/desktop/CUNY Internship/Terra"
db_directory <- "C:/users/owner/desktop/CUNY Internship/Terra"

# Load the dataframe of Terra Workspaces and their metadata
all_spaces <- read.csv(file = file.path(db_directory, "workspaces.csv"), header=T)

# Load the dataframe of Terra Workflows and their metadata
all_flows <- read.csv(file = file.path(db_directory, "workflows.csv"), header=T)

# Load the dataframe of Terra sample data and their metadata
all_data <- read.csv(file = file.path(db_directory, "sampledata.csv"), header=T)
```

```{r}
dir <- system.file("extdata", package = "AnVILWorkflow")
all_spaces <- read_csv(file.path(dir, "AnVILworkspaces.csv"))
all_flows <- read_csv(file.path(dir, "AnVILworkflows.csv"))
all_data <- read_csv(file.path(dir, "sampledata.csv"))
```


## Loading Base Functions

Next, we can load the base functions required to run search queries.
```{r}
# Primary subfunction for returning table of search result hits
primary_df_query <- function(search_param, data, columns){
  rows <- c()
  for (row in 1:nrow(data)){
    if (T %in% grepl(search_param, as.character(data[row,columns]), ignore.case=T)){
      # add row to list of rows
      rows <- c(rows, row)
    }
  }
  primary_df_hits <- data[rows,]
  
  return(primary_df_hits)
}

# Secondary subfunction for returning table of search result hits
secondary_df_query <- function(search_param, flowdata, spacedata, columns){
  rows <- c()
  for (row in 1:nrow(spacedata)){
    if (T %in% grepl(search_param, as.character(spacedata[row,columns]), ignore.case=T)){
      # add row to list of rows
      rows <- c(rows, row)
    }
  }
  secondary_df_hits <- flowdata[which(flowdata$workspace_key %in% spacedata$workspace_key[rows]),]
  
  return(secondary_df_hits)
}

# Search function for Workspaces
workspace_search <- function(search_param, data=all_spaces, columns=(1:ncol(data))){
  search_results <- primary_df_query(search_param, data, columns)
  if (!nrow(search_results)){
    print("No workspaces matched your search criteria.")
  }else{
    return(search_results)
  }
}

# Search function for Workflows
workflow_search <- function(search_param, flowdata=all_flows, spacedata=all_spaces, 
                            flowcolumns=(1:ncol(flowdata)), spacecolumns=(1:ncol(spacedata)),
                            join="u"){
  # Query the workflows dataframe
  p_search_results <- primary_df_query(search_param, flowdata, flowcolumns)
  # Query the workspaces dataframe
  s_search_results <- secondary_df_query(search_param, flowdata, spacedata, spacecolumns)
  
  # Compile query results
  if ((nrow(p_search_results) > 0) & (nrow(s_search_results) > 0)){
    if (join=="i"){
      search_results <- p_search_results[which(p_search_results$name_key %in% s_search_results$name_key),]
    }else{
      search_results <- rbind(s_search_results, p_search_results[which(!(p_search_results$name_key %in% s_search_results$name_key)),])
    }
  }else if(nrow(p_search_results) > 0){
    search_results <- p_search_results
  }else{
    search_results <- s_search_results
  }
  
  # Return results if applicable
  if (!nrow(search_results)){
    print("No workflows matched your search criteria.")
  }else{
    return(search_results)
  }
}

# Search function for Sample Data
sampledata_search <- function(search_param, sampledata=all_data, spacedata=all_spaces, 
                            samplecolumns=(1:ncol(sampledata)), 
                            spacecolumns=(1:ncol(spacedata)), join="u"){
  # Query the workflows dataframe
  p_search_results <- primary_df_query(search_param, sampledata, samplecolumns)
  # Query the workspaces dataframe
  s_search_results <- secondary_df_query(search_param, sampledata, spacedata, spacecolumns)
  
  # Compile query results
  if ((nrow(p_search_results) > 0) & (nrow(s_search_results) > 0)){
    if (join=="i"){
      search_results <- p_search_results[which(p_search_results$name_key %in% s_search_results$name_key),]
    }else{
      search_results <- rbind(s_search_results, p_search_results[which(!(p_search_results$name_key %in% s_search_results$name_key)),])
    }
  }else if(nrow(p_search_results) > 0){
    search_results <- p_search_results
  }else{
    search_results <- s_search_results
  }
  
  # Return results if applicable
  if (!nrow(search_results)){
    print("No sample data matched your search criteria.")
  }else{
    return(search_results)
  }
}
```


## Applying the search functions

Now we can apply the query functions. Some examples are provided to demonstrate
the applications of each main search function. It is worth noting that results
from workflow and sample data searches also query workspaces, and results from
these queries are additive. This means that if you specify a workflow search 
column, you must assign `spacecolumns` to a zero value (0 or "").
```{r export_curated_table, eval=FALSE}
## WORKSPACE QUERIES ##

# Entering a workspace search parameter only
workspace_search("lymphoma") # returns 2 workspace results

# Entering a workspace query that returns no results
workspace_search("lung cancer")

# Searching for matches in a specific column of workspace data using column name
workspace_search("usa", columns="cohort_country") # returns 68 workspace results
workspace_search("usa") # comparatively, returns 105 workspace results


## WORKFLOW QUERIES ##

# Entering a workflow search parameter only
workflow_search("malaria") # returns 35 results

# Searching for matches in a specific column of workflow data using column number
workflow_search("malaria", flowcolumns=1, spacecolumns=0) # returns 3 results

# Entering a workflow query that returns no results
workflow_search("malaria", flowcolumns=6, spacecolumns="")

# Searching for matches in a specific column of workspace data using column numbers
workflow_search("malaria", flowcolumns=0, spacecolumns=5:8) # returns 35 results

# Searching for matches using both workflow and workspace column specifications
workflow_search("resistance", flowcolumns=c("name", "namespace"), 
                spacecolumns="description") # returns 25 results

# Searching for matches using both workflow and workspace column specifications
workflow_search("resistance", flowcolumns=c("name", "namespace"), 
                spacecolumns="description", join="i") # returns 2 results


## SAMPLE DATA QUERIES ##

# Entering a sample data search parameter only
sampledata_search("lymph node") # returns 7 results

# Entering a sample data query that returns no results
sampledata_search("lung cancer")

# Searching for matches in a specific column of sample data using column names
sampledata_search("female", samplecolumns="GENDER", spacecolumns="") # returns 37 results

```